Thesis: draft

### 2021-04-27 draft.tex
latest update

\renewcommand{\abstractname}{Abstract}
\selectlanguage{english}
\begin{abstract}
\addcontentsline{toc}{chapter}{Abstract}
The past decade, the rise of social media has become undeniable. More and more people are getting engaged on some type of social media platform and hence, it is probably not wrong to say that nowadays most of us have, beside our real life, a virtual life on social media. This rapidly increased engagement on social media undoubtedly has many advantages. However, every upside has a downside and social media is no different. This thesis deals with the impact of social media on the formation and evolution of opinions in society.\\ 
Social media allows for the spread of information at a faster pace and at an unprecedented scale. However, people are still bound by time- and cognitive constraints. In order to handle the information overflow, social media companies design filtering algorithms that can fine-tune and individualize the information someone is exposed to. This filtering and individualizing of information may have some potentially dangerous downfalls in the sense that people might get trapped in ``echo chambers'' that confirm their own beliefs. It is important to know the effects of such algorithmic personalization on the appearance and formation of phenomena such as polarization, spread of extremism, formation of echo chambers, etc.\\
People are, however, more than sheep that thoughtlessly follow their shepherd (contrary to what some books might lead us to believe). They are capable of critical thinking and may be hesitant to change their current opinion. This aspect of potential stubbornness should not be ignored.\\
In this thesis toy models are designed to investigate the interplay between network structure, filtering algorithms and individual resistance to change in the evolution of two competing opinions.\\
We will investigate whether network properties such as community structure and clustering coefficient may enhance or hamper the formation of echo chambers. We will use some basic filtering algorithms to determine their possible effects on the appearance of echo chambers and we will investigate the impact of individual stubbornness. \\
Beside that, we will also pay some attention to the possible ways of distributing the opinions and the stubborn people and how this may effect the opinion dynamics in the system. \\
Finally, we will test the model on real-world networks. Some question that may arise are: Do the results agree with those obtained from the network models? What are the similarities or differences in the results? What are the possible underlying reason/structures for these similarities/differences?\\

@@

The past decade social media has become ever more important in our daily lives. Nowadays most of us have, beside a real-world life, a virtual life on social media. This increased engagement on social media may have a huge impact on the formation of opinions, not only on the individual scale, but also on a broader, societal scale. \\
It is known that individuals not only form their opinions through self-reflection, but also through interactions with other people and with their surroundings \cite{Perra2019}. The broad range of interactions people undergo on social media with people from all over the world can thus not be underestimated in the process of opinion formation. The understanding of the role of social media on the emergence of polarization and extremism in society is of uttermost importance.\\
\newline
It is also known that some people change their opinions more easily than others, that is, once in a while you encounter somebody that is really stubborn and persistent towards its own opinion. One of the aims of this thesis is to design toy models of opinion dynamics with stubborn actors on theoretical and real-world social networks to analyze the interplay between individual resistance to change and the network structure in the evolution of two competing opinions.\\
\newline
One important feature of opinion dynamics on social media is the use of algorithmic personalization. Social media allow for simultaneous or asynchronous interactions between people without geographical constraints and thus allow for the spread of information at a faster pace and at an unprecedented scale \cite{Perra2019}. People are, however, still bound by temporal and cognitive constraints. In order to assure a more pleasant/convenient experience on their platforms, social media companies use algorithms to order/filter the posts on an individual's time-line according to what might be relevant to that individual \cite{Perra2019}. It is of great importance to know the possible effects of these filtering algorithms on the evolution of opinions in the population.\\  
\newline
Opinion dynamics models have two important layers. On one hand, we introduce social networks to describe the underlying structure of social interactions; on the other hand, we need appropriate opinion dynamics/formation models to reproduce the opinion dynamics and opinion formation processes found in real life.\\
\newline
Section \ref{complNet} gives an introduction to the first layer, namely that of complex networks. The second layer, the opinion dynamics layer, is introduced in Section \ref{Opinion}. Section \ref{statPhys} introduces concepts such as statistical and social physics. Finally, in Section \ref{goal}, the hypotheses and goals of this thesis are discussed.

@@

As stated before, the formation of an individual's opinion is the result of the interplay of many factors such as self-reflection, peer pressure, the individual's personality (eg. stubbornness), the information someone is exposed to, etc.  Opinion dynamics models should try to capture these complex processes in a simplified way. Several models have been developed ranging from simple binary models to more complex, continuous approaches \cite{Sirbu2016}.\\
The basic idea of all opinion dynamics models is that the nodes or agents in a social network have a variable that represents their opinion and that is updated according to some predefined rules. These models are obviously a simplification of real-world opinion dynamics. It is however shown that they do display a lot of aspects of real opinion formation such as agreement, transitions between order (consensus) and disorder (fragmentation), polarization, formation of echo chambers (clusters of people with the same opinion), etc \cite{Sirbu2016}. \\
\newline
The opinion of the agents in the model can be either discrete or continuous. This thesis deals with models where each agent can have one of the two opinions A or B and thus only deals with the case of discrete opinions. This might come across as a huge simplification of the real-world complexity of opinion formation, but also in real life people often have to choose between two competing opinions (eg. republicans or democrats, cat or dog, renting or buying a house,...). \\
\newline
The rules that determine how an agent updates his or her opinion can include many different aspects. They can, for example, be as simple as a majority rule (i.e. the majority model: if the majority of your neighbors have a certain opinion, you adopt that opinion as well) or can include a more probabilistic way of updating. A small overview of possible opinion dynamics models is given in Chapter \ref{chap3}, Section \ref{Bin} and Section \ref{Con}. \\
One can also include a resistance parameter that determines the hesitation of an agent to change to a new opinion and/or a parameter that determines the influence of an agent on others. Furthermore, the agents can have a parameter that determines whether they are active or not. The introduction of stubborn people will be discussed in Chapter \ref{chap3}, Section \ref{stubb}; the activation mechanism of nodes is discussed in Chapter \ref{chap3}, Section \ref{actMech}.\\
\newline
Since this thesis deals with the particular case of opinion dynamics on on-line social platforms, it is important to define/set up different filtering algorithms. The filtering algorithms that real social media companies use, are corporate secrets, but we know that there are three main principles of content curation: popularity, semantic and collaborative filtering \cite{Perra2019}. Popularity filtering refers to the practice of promoting content that is popular across the platform; semantic filtering means that post similar to previous consumed posts are recommended and collaborative filtering suggests posts that are similar to the ones our friends consume \cite{Perra2019}. In Chapter \ref{chap3}, Section \ref{filter} a detailed description of the filtering algorithms used in  this thesis will be given.

@@

In 1998, Duncan Watts and Steve Strogatz constructed a network model that generates random networks with small world properties, such as high clustering and short average distances. Their model achieves this by interpolating between a regular ring lattice and a random network \cite{Watts1998}. The construction goes as follows \cite{Watts1998}
\begin{enumerate}
	\item Start with a regular ring lattice with $N$ nodes, each connected to $K = 2m$ neighbors ($m$ on each side).
	\item For each node $i$, rewire each edge that connects $i$ with its $K/2$ rightmost neighbors with a probability $\beta$. Rewiring is done in such a way that self-loops and edge duplication are avoided.
\end{enumerate}
The probability $\beta$ allows us to tune the graph between a regular lattice ($\beta = 0$) and a random network ($\beta = 1$) \cite{Watts1998}. Usually, the following condition is required for the number of neighbors $K$: $N \gg K \gg \ln{N} \gg 1$. The first inequality assures sparse networks, whereas the second inequality guarantees that a random graph will be connected (thus the network is not so sparse that it becomes disconnected) \cite{Watts1998}.\\
The regular ring lattice ($\beta=0$) has an average clustering coefficient $\left<C(0)\right>$ and an average path length $\left<d(0)\right>$ equal to \cite{Watts1998}
\begin{align}\label{C0}
	\left<C(0)\right> = \frac{3(K-2)}{4(K-1)} \approx \frac{3}{4} \ \text{for}\ K \gg 1 \ , && \left<d(0)\right> \approx \frac{N}{2K} \ .
\end{align}
The derivations of these formulas can be found in Appendix \ref{WS}, Section \ref{ringclus}. The regular ring lattice thus has a high clustering coefficient and an average path length that grows linearly with the system size $N$. \\
The random network ($\beta = 1$) on the other hand has a low clustering coefficient $\left<C(1)\right>$ and a small average path length $\left<d(1)\right>$ (see Eq. (\ref{C}) and Eq. (\ref{d})) \cite{Watts1998}
\begin{align}
	\left<C(1)\right> \approx \frac{K}{N} \ll 1 \ , && \left<d(1)\right> \approx \frac{\ln{N}}{\ln{K}} \ .
\end{align}
Watts and Strogatz then showed by numerical simulation that the average distance for the networks with $0 < \beta < 1$ is comparable with the average distance of a random graph (which is small), even for small $\beta$ \cite{Newman2000}. The clustering coefficient is approximately given by   \cite{Barrat1999}
\begin{equation}
	\left<C\right> \approx \left<C(0)\right>(1-\beta)^3 \ ,
\end{equation} 
where $\left<C(0)\right>$ is given by Equation (\ref{C0}). This can be seen in the following way
\begin{itemize}
	\item The probability that none of the three edges of the original triangles in the regular lattice are rewired is $(1-\beta)^3$.
	\item The probability that edges are rewired back to each other is negligible.
\end{itemize}
For rather small values of $\beta$, the clustering coefficient remains high. The model of Watts and Strogatz is thus able to generate random networks with a high clustering coefficient and short average distances. The Watts-Strogatz network has a homogeneous, bell shaped degree distribution with an average degree equal to $\left<k\right> = K$.

@@

In continuous opinion dynamics models, the opinion variable of the agents in the network is no longer discrete, in contrast to binary or discrete models, but takes real values (often in the interval $\interval{0}{1}$ or $\interval{-1}{1}$).\\
In continuous opinion dynamics models, some concepts that are valid for the discrete models no longer apply (eg. the concept of a majority/minority opinion or the concept of equality of opinions); hence, they require a bit of a different framework \cite{Castellano2009}. \\
The basis of most continuous models is a set of $N$ nodes or agents, each having an opinion $x_i$ represented by a real number that takes values in some interval \cite{Castellano2009}. In contrast to discrete/binary models, all agents usually start with  different opinions \cite{Castellano2009}. The possible outcomes are more complex than in the discrete models and often has opinion clusters emerging in the final states (where opinion clusters refer to clusters of nodes with the same opinion) \cite{Castellano2009}. The number of opinion clusters can be one (consensus), two (polarization) or more (fragmentation) \cite{Castellano2009}. \\
Several models have been proposed, we will only discuss a fraction of them.

@@

It is important to note that what somebody sees on his/her time-line is the consequence of the interplay between the filtering algorithm and the posted opinions of his/her neighbors. Every user has a ``hidden time-line $L_i(t)$'' which contains the actual time-line of the user at the moment when he/she was active for the last time ($t_{\text{last}}$) plus all the posts of the neighbors from that last time he/she was active till the current time he/she becomes active again ($t$) \cite{Perra2019}. At the time $t$ that the user becomes active again, the filtering algorithm selects $Q$ post from this ``hidden time-line $L_i(t)$'' and shows them on the actual time-line of the user; these $Q$ items are thus the opinions that the user is exposed to \cite{Perra2019} (see also Section \ref{filter} for more detailed information on the filtering algorithms).

@@

It is well known that in real life, you can find people that are really persistent towards their own opinion and that are not open to opinions of others. They are convinced that they carry the truth with them and that other people, with different world-views, are not capable of giving them any new and useful insights. They are close-minded in contrast to people with an open-mind that are willing to learn from others and to reconsider their beliefs. Of course this is put a bit simplistic and there are many shades of gray between this black and white statement. However, there are groups in society, often with more extreme opinions, that are know to be really inflexible towards changing that (extreme) opinion (note here that being stubborn and being extreme doesn't need to go hand in hand, you can obviously be stubborn without being extreme and vice versa; however, it is often widely accepted that there is some kind of correlation/link between people with extreme opinions and being more persistent towards these opinions compared to people with more moderate beliefs; this idea of correlating extreme people with a more stubborn character is also often used in models, see for example \cite{Deffuant2006}\cite{Deffuant2002}\cite{Turner}) (also note that having some resistance to change your opinion to any other opinion doesn't necessarily need to be a bad thing, it can be seen as a form of critical thinking and as a form of taking things with a grain of salt).\\ 
\newline
Knowing the impact of such stubborn groups on the opinions of others is of great importance, especially with the rise of social media which makes the reach of such extreme and persistent groups a lot bigger. Before the era of social media, one could maybe argue that these extreme and stubborn groups are often isolated and therefore wouldn't reach too many people. So their impact could only be of a minor nature (note that this is only a statement that could have been given, there are of course plenty of examples throughout human history that contradict this argument; to only mention a few: Nazism in the 1930s and 1940s, the Ku Klux Klan in the late 19th and begin 20th century, the rise of communism in the 20th century, etc, all examples of groups that can be considered extreme and that have had a wide reach and a significant impact on society). Today, this possible argument is no longer valid; social media enables the contact with all sorts of beliefs and all sorts of people with almost no (time and spatial) constraints. Extreme and stubborn groups that may have been isolated before no longer need to be so and this may have huge, and largely unknown, impacts. \\
\newline
What is the impact of having stubborn/extreme people in your society? Does it matter if the stubborn people are grouped together or if they are distributed evenly throughout your system? Are there ways to reduce the possible impacts of stubbornness and/or extremism? Can these stubborn groups lead to polarization or even overtake the opinions in the system? Or is introducing some kind of stubbornness, on the other hand, a protective factor towards extremism? To answer these questions, several models have been proposed to include some kind of stubbornness/resistance in the system under study.

@@

	\item If $0 < r_i < 1$, the node has some resistance but can still be convinced to switch to the other opinion, even though he/she will be less likely to do so. A uniform random number $f$ between 0 and 1 is drawn. If $f > r_i$, the node will update his/her opinion; if $f \leqslant r_i$, the node is not convinced and keeps his/her own opinion.

@@

The previous two chapters, Chapter \ref{chap2} and \ref{chap3}, gave a broad introduction/overview of networks and opinion dynamics models respectively. In this chapter we will summarize the specific networks and opinion dynamics models that are used in this thesis. \\
Section \ref{networkmodel} gives an overview of the network models that are used in this thesis. In chapter \ref{chap2}, Section \ref{netModel} we already discussed their main properties; here we will mainly discuss why we use these networks. Which of their properties are of interest in this thesis? What are there benefits and possible shortcomings regarding to our goals?\\
In section \ref{realnetwork}, we will discuss the real-world networks used in this thesis. The reasons of choosing these specific networks will be explained, together with there possible strengths and weaknesses.\\ % + explain where they come from etc?
Finally, Section \ref{opinionmodel} discusses the opinion dynamics model that is implemented in this thesis. Again we will give some focus on possible advantages and disadvantages of the model.

@@

This model is a very simple one, since it only allows for two different opinions in the system. The designed filtering algorithms are very simple and intuitive as well. It also introduces some randomness by using the probabilistic majority model instead of the classic majority model. This might make the model a bit more realistic. This simpleness combined with the touch of randomness to give the model a more realistic vibe are definitely some strong points of using this model compared to other, more complex models. A simple model is more tractable and allows for a better understanding of what is going on in the dynamics. However, this might be a downside as well. Real-world opinion dynamics are, probably, far from simple. People don't update their opinion according to a simple probabilistic opinion dynamics model. Beside that, the cases where people have to choose between only two opinions are rather limited. Filtering algorithms, used by real social media companies, are unfortunately not publicly shared. We can however assume that they are not as simple as the ones we used in this thesis. Trying to model these far more complex systems by our very simple model, might lead to the fact that a lot of the dynamics might be missed or not well captured by our model. This downside of trying to model real-world social behavior with simple, tractable models is really a call for the urgent need for more real-world data to which the results obtained from simple models can be compared and which can help make the models capable of generating more realistic results. \\

@@

Note that for an initial $50/50$ opinion distribution none of the network models, nor the filtering algorithms could break this status quo (see Figure \ref{ev_op_50_50}). Despite this preserving of the balanced opinion distribution, the interplay of filtering algorithms and network structure can potentially shift the initial randomly distributed opinions to a more clustered opinion distribution and thus lead to polarization and formation of echo chambers.

@@

This enhancing of echo chambers around the minority opinion in case of networks with a high clustering coefficient is remarkable, especially in the case of the PR filtering algorithm. Remember from Figure \ref{ev_op_20_80} that in case of the PR method the minority opinion 0 decreases to less than $1 \%$ of the entire population for the high clustering networks. Even though the total fraction of the minority opinion strongly decreases, the high clustering forces the remaining nodes with this minority opinion to group together and form opinion clusters. The normalized value of the formation of echo chambers around the majority opinion reaches values approximately equal to $F_N(\left<P_1^{\text{nn}}\right> = 1) = 1$ when using the REC filtering algorithm. This means that the average fraction of nodes with all their neighbors carrying the majority opinion 1 remains the same before and after the time evolution. In case of the PR method this value is even slightly higher than 1, indicating a small increase in the number of majority opinion echo chambers after the time evolution.

@@

\subsection{Including a stubbornness parameter}
\label{stubbpar}
Here, we review the results of introducing stubborn people by including a stubbornness parameter $r_i$. We look at two slightly different case
\begin{itemize}
	\item Making a fraction of the nodes (going from none of the nodes to all the nodes) completely stubborn ($r_i = 1$),
	\item Making all nodes partly stubborn (going from not stubborn at all, $r = 0$ to completely stubborn, $r = 1$).
\end{itemize}
For both cases we will look at both the evolution of the opinions and the formation of echo chambers and we will do this for two different initial conditions: a balanced initial opinion distribution ($50/50$) and an unbalanced initial opinion distribution ($20/80$). \\
Once again, the opinion dynamics model described in Chapter \ref{chap4}, Section \ref{opinionmodel} is implemented with an activation probability equal to $p_{\text{act}} = 0.1$ and a maximum length of the actual time-line $S = 20$.

@@

  	\includegraphics[width=8cm, height=7cm]{fraction_of_opinions_ER_frac_compl_stubb_PR_20-80_8x7.png}
    \caption{ER}
  	\includegraphics[width=8cm, height=7cm]{fraction_of_opinions_WS_frac_compl_stubb_PR_20-80_8x7.png}
    \caption{WS}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM_frac_compl_stubb_PR_20-80_8x7.png}
    \caption{SBM}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM-WS_frac_compl_stubb_PR_20-80_8x7.png}
    \caption{SBM-WS}

>>>>

\renewcommand{\abstractname}{Abstract}
\selectlanguage{english}
\begin{abstract}
\addcontentsline{toc}{chapter}{Abstract}
The past decade, the rise of social media has become undeniable. More and more people are getting engaged on some type of social media platform and hence, it is probably not wrong to say that nowadays most of us have, beside our real life, a virtual life on social media. This rapidly increased engagement on social media undoubtedly has many advantages. However, every upside has a downside and social media is no different. This thesis deals with the impact of social media on the formation and evolution of opinions in society.\\ 
Social media allows for the spread of information at a faster pace and at an unprecedented scale. However, people are still bound by time- and cognitive constraints. In order to handle the information overflow, social media companies design filtering algorithms that can fine-tune and individualize the information someone is exposed to. This filtering and individualizing of information may have some potentially dangerous downsides in the sense that people might get trapped in ``echo chambers'' that confirm their own beliefs. It is important to know the effects of such algorithmic personalization on the appearance and formation of phenomena such as polarization, spread of extremism, formation of echo chambers, etc.\\
People are, however, more than sheep that thoughtlessly follow their shepherd (contrary to what some books might lead us to believe). They are capable of critical thinking and may be hesitant to change their current opinion. This aspect of potential stubbornness should not be ignored.\\
In this thesis toy models are designed to investigate the interplay between network structure, filtering algorithms and individual resistance to change in the evolution of two competing opinions.\\
We will investigate whether network properties such as community structure and clustering coefficient may enhance or hamper the formation of echo chambers. We will use some basic filtering algorithms to determine their possible effects on the appearance of echo chambers and we will investigate the impact of individual stubbornness. \\
Beside that, we will also pay some attention to the possible ways of distributing the opinions and the stubborn people and how this may effect the opinion dynamics in the system. \\
Finally, we will test the model on real-world networks. Some question that may arise are: Do the results agree with those obtained from the network models? What are the similarities or differences in the results? What are the possible underlying reason/structures for these similarities/differences?\\

@@

The past decade social media has become ever more important in our daily lives. Nowadays most of us have, beside a real-world life, a virtual life on social media. This increased engagement on social media may have a huge impact on the formation of opinions, not only on the individual scale, but also on a broader, societal scale. \\
It is known that individuals not only form their opinions through self-reflection, but also through interactions with other people and with their surroundings \cite{Perra2019}. The broad range of interactions people undergo on social media with people from all over the world can thus not be underestimated in the process of opinion formation. The understanding of the role of social media on the emergence of polarization and extremism in society is of uttermost importance.\\
\newline
It is also known that some people change their opinions more easily than others, that is, once in a while you encounter somebody that is really stubborn and persistent towards its own opinion. One of the aims of this thesis is to design toy models of opinion dynamics with stubborn actors on theoretical and real-world social networks to analyze the interplay between individual resistance to change and the network structure in the evolution of two competing opinions.\\
\newline
One important feature of opinion dynamics on social media is the use of algorithmic personalization. Social media allow for simultaneous or asynchronous interactions between people without geographical constraints and thus allow for the spread of information at a faster pace and at an unprecedented scale \cite{Perra2019}. People are, however, still bound by temporal and cognitive constraints. In order to assure a more pleasant/convenient experience on their platforms, social media companies use algorithms to order/filter the posts on an individual's time-line according to what might be relevant to that individual \cite{Perra2019}. It is of great importance to know the possible effects of these filtering algorithms on the evolution of opinions in the population.\\  
\newline
Opinion dynamics models have two important layers. On one hand, we introduce social networks to describe the underlying structure of social interactions; on the other hand, we need appropriate opinion dynamics/formation models to reproduce the opinion dynamics and opinion formation processes found in real life.\\
\newline
Section \ref{complNet} gives an introduction to the first layer, namely that of complex networks. The second layer, the opinion dynamics layer, is introduced in Section \ref{Opinion}. Section \ref{statPhys} introduces concepts such as statistical and social physics. Finally, in Section \ref{goal}, the hypotheses and goals of this thesis are summarized and discussed.

@@

As stated before, the formation of an individual's opinion is the result of the interplay of many factors such as self-reflection, peer pressure, the individual's personality (eg. stubbornness), the information someone is exposed to, etc.  Opinion dynamics models should try to capture these complex processes in a simplified way. Several models have been developed ranging from simple binary models to more complex, continuous approaches \cite{Sirbu2016}.\\
The basic idea of all opinion dynamics models is that the nodes or agents in a social network have a variable that represents their opinion and that is updated according to some predefined rules. These models are obviously a simplification of real-world opinion dynamics. It is however shown that they do display a lot of aspects of real opinion formation such as agreement, transitions between order (consensus) and disorder (fragmentation), polarization, formation of echo chambers (clusters of people with the same opinion), etc \cite{Sirbu2016}. \\
\newline
The opinion of the agents in the model can be either discrete or continuous. This thesis deals with models where each agent can have one of the two opinions A or B and thus only deals with the case of discrete opinions. This might come across as a huge simplification of the real-world complexity of opinion formation, but also in real life people often have to choose between two competing opinions (eg. republicans or democrats, cat or dog, renting or buying a house,...). \\
\newline
The rules that determine how an agent updates his or her opinion can include many different aspects. They can, for example, be as simple as a majority rule (i.e. the majority model: if the majority of your neighbors have a certain opinion, you adopt that opinion as well) or can include a more probabilistic way of updating. A small overview of possible opinion dynamics models is given in Chapter \ref{chap3}, Section \ref{Bin} and Section \ref{Con}. \\
One can also include a resistance parameter that determines the hesitation of an agent to change to a new opinion and/or a parameter that determines the influence of an agent on others. Furthermore, the agents can have a parameter that determines whether they are active or not. The introduction of stubborn people will be discussed in Chapter \ref{chap3}, Section \ref{stubb}; the activation mechanism of nodes is discussed in Chapter \ref{chap3}, Section \ref{actMech}.\\
\newpage
\noindent
Since this thesis deals with the particular case of opinion dynamics on on-line social platforms, it is important to define/set up different filtering algorithms. The filtering algorithms that real social media companies use, are corporate secrets, but we know that there are three main principles of content curation: popularity, semantic and collaborative filtering \cite{Perra2019}. Popularity filtering refers to the practice of promoting content that is popular across the platform; semantic filtering means that post similar to previous consumed posts are recommended and collaborative filtering suggests posts that are similar to the ones our friends consume \cite{Perra2019}. In Chapter \ref{chap3}, Section \ref{filter} a detailed description of the filtering algorithms used in  this thesis will be given.

@@

In 1998, Duncan Watts and Steve Strogatz constructed a network model that generates random networks with small world properties, such as high clustering and short average distances. Their model achieves this by interpolating between a regular ring lattice and a random network \cite{Watts1998}. The construction goes as follows \cite{Watts1998}
\begin{enumerate}
	\item Start with a regular ring lattice with $N$ nodes, each connected to $K = 2m$ neighbors ($m$ on each side).
	\item For each node $i$, rewire each edge that connects $i$ with its $K/2$ rightmost neighbors with a probability $\beta$. Rewiring is done in such a way that self-loops and edge duplication are avoided.
\end{enumerate}
The probability $\beta$ allows us to tune the graph between a regular lattice ($\beta = 0$) and a random network ($\beta = 1$) \cite{Watts1998}. Usually, the following condition is required for the number of neighbors $K$: $N \gg K \gg \ln{N} \gg 1$. The first inequality assures sparse networks, whereas the second inequality guarantees that a random graph will be connected (thus the network is not so sparse that it becomes disconnected) \cite{Watts1998}.\\
The regular ring lattice ($\beta=0$) has an average clustering coefficient $\left<C(0)\right>$ and an average path length $\left<d(0)\right>$ equal to \cite{Watts1998}
\begin{align}\label{C0}
	\left<C(0)\right> = \frac{3(K-2)}{4(K-1)} \approx \frac{3}{4} \ \text{for}\ K \gg 1 \ , && \left<d(0)\right> \approx \frac{N}{2K} \ .
\end{align}
The derivations of these formulas can be found in Appendix \ref{WS}, Section \ref{ringclus}. The regular ring lattice thus has a high clustering coefficient and an average path length that grows linearly with the system size $N$. \\
The random network ($\beta = 1$) on the other hand has a low clustering coefficient $\left<C(1)\right>$ and a small average path length $\left<d(1)\right>$ (see Eq. (\ref{C}) and Eq. (\ref{d})) \cite{Watts1998}
\begin{align}
	\left<C(1)\right> \approx \frac{K}{N} \ll 1 \ , && \left<d(1)\right> \approx \frac{\ln{N}}{\ln{K}} \ .
\end{align}
Watts and Strogatz then showed by numerical simulation that the average distance for the networks with $0 < \beta < 1$ is comparable with the average distance of a random graph (which is small), even for small $\beta$ \cite{Newman2000}. The clustering coefficient is approximately given by   \cite{Barrat1999}
\begin{equation}
	\left<C\right> \approx \left<C(0)\right>(1-\beta)^3 \ ,
\end{equation} 
where $\left<C(0)\right>$ is given by Equation (\ref{C0}). This can be seen in the following way
\begin{itemize}
	\item The probability that none of the three edges of the original triangles in the regular lattice are rewired is $(1-\beta)^3$.
	\item The probability that edges are rewired back to each other is negligible.
\end{itemize}
For rather small values of $\beta$, the clustering coefficient remains high. The model of Watts and Strogatz is thus able to generate random networks with a high clustering coefficient and short average distances. The Watts-Strogatz network has a homogeneous, bell shaped degree distribution with an average degree equal to $\left<k\right> = K$ \cite{Barrat1999}.

@@

In continuous opinion dynamics models, the opinion variable of the agents in the network is no longer discrete, in contrast to binary or discrete models, but takes real values (often in the interval $\interval{0}{1}$ or $\interval{-1}{1}$).\\
In continuous opinion dynamics models, some concepts that are valid for the discrete models no longer apply (eg. the concept of a majority/minority opinion or the concept of equality of opinions); hence, they require a bit of a different framework \cite{Castellano2009}. \\
The basis of most continuous models is a set of $N$ nodes or agents, each having an opinion $x_i$ represented by a real number that takes values in some interval \cite{Castellano2009}. In contrast to discrete/binary models, all agents usually start with  different opinions \cite{Castellano2009}. The possible outcomes are more complex than in the discrete models and often have opinion clusters emerging in the final states (where opinion clusters refer to clusters of nodes with the same opinion) \cite{Castellano2009}. The number of opinion clusters can be one (consensus), two (polarization) or more (fragmentation) \cite{Castellano2009}. \\
Several models have been proposed, we will only discuss a fraction of them.

@@

It is important to note that what somebody sees on his/her time-line is the consequence of the interplay between the filtering algorithm and the posted opinions of his/her neighbors. Every user has a ``hidden time-line $L_i(t)$'' which contains the actual time-line of the user at the moment when he/she was active for the last time ($t_{\text{last}}$) plus all the posts of the neighbors from that last time he/she was active till the current time he/she becomes active again ($t$) \cite{Perra2019}. At the time $t$ that the user becomes active again, the filtering algorithm selects $S$ post from this ``hidden time-line $L_i(t)$'' and shows them on the actual time-line of the user; these $S$ items are thus the opinions that the user is exposed to \cite{Perra2019} (see also Section \ref{filter} for more detailed information on the filtering algorithms).

@@

It is well known that in real life, you can find people that are really persistent towards their own opinion and that are not open to opinions of others. They are convinced that they carry the truth with them and that other people, with different world-views, are not capable of giving them any new and useful insights. They are close-minded in contrast to people with an open-mind that are willing to learn from others and to reconsider their beliefs. Of course this is put a bit simplistic and there are many shades of gray between this black and white statement. However, there are groups in society, often with more extreme opinions, that are know to be really inflexible towards changing that (extreme) opinion (note that being stubborn and being extreme doesn't need to go hand in hand, you can obviously be stubborn without being extreme and vice versa; however, it is often widely accepted that there is some kind of correlation/link between people with extreme opinions and being more persistent towards these opinions compared to people with more moderate beliefs; this idea of correlating extreme people with a more stubborn character is also often used in models, see for example \cite{Deffuant2006}\cite{Deffuant2002}\cite{Turner}) (also note that having some resistance to change your opinion to any other opinion doesn't necessarily need to be a bad thing, it can be seen as a form of critical thinking and as a form of taking things with a grain of salt).\\ 
\newline
Knowing the impact of such stubborn groups on the opinions of others is of great importance, especially with the rise of social media which makes the reach of more extreme and persistent groups a lot bigger. Before the era of social media, one could maybe argue that these extreme and stubborn groups are often isolated and therefore wouldn't reach too many people. So their impact could only be of a minor nature (note that this is only a statement that could have been given, there are of course plenty of examples throughout human history that contradict this argument; to only mention a few: Nazism in the 1930s and 1940s, the Ku Klux Klan in the late 19th and begin 20th century, the rise of communism in the 20th century, etc, all examples of groups that can be considered extreme and that have had a (relatively) wide reach and a significant impact on society). Today, this possible argument is no longer valid; social media enables the contact with all sorts of beliefs and all sorts of people with almost no (time and spatial) constraints. Extreme and stubborn groups that may have been isolated before no longer need to be so and this may have huge, and largely unknown, impacts. \\
\newline
What is the impact of having stubborn/extreme people in your society? Does it matter if the stubborn people are grouped together or if they are distributed evenly throughout your system? Are there ways to reduce the possible impacts of stubbornness and/or extremism? Can these stubborn groups lead to polarization or even overtake the opinions in the system? Or is introducing some kind of stubbornness, on the other hand, a protective factor towards extremism/polarization? To answer these questions, several models have been proposed to include some kind of stubbornness/resistance in the system under study.

@@

	\item If $0 < r_i < 1$, the node has some resistance but can still be convinced to switch to the other opinion, even though he/she will be less likely to do so. An uniform random number $f$ between 0 and 1 is drawn. If $f > r_i$, the node will update his/her opinion; if $f \leqslant r_i$, the node is not convinced and keeps his/her own opinion.

@@

The previous two chapters, Chapter \ref{chap2} and \ref{chap3}, gave a broad introduction/overview of networks and opinion dynamics models respectively. In this chapter we will summarize the specific networks and opinion dynamics models that are used in this thesis. \\
Section \ref{networkmodel} gives an overview of the network models that are used in this thesis. In chapter \ref{chap2}, Section \ref{netModel} we already discussed their main properties; here we will mainly discuss why we use these networks. Which of their properties are of interest in this thesis? What are there benefits and possible shortcomings regarding to our goals?\\
In section \ref{realnetwork}, we will discuss the real-world networks used in this thesis. The reasons of choosing these specific networks will be explained, together with there possible strengths and weaknesses.\\ % + explain where they come from etc?
Finally, Section \ref{opinionmodel} summarizes and discusses the opinion dynamics model that is implemented in this thesis. Again we will give some focus on possible advantages and disadvantages of the model.

@@

This model is a very simple one, since it only allows for two different opinions in the system. The designed filtering algorithms are very simple and intuitive as well. Some randomness is introduced by using the probabilistic majority model instead of the classic majority model. This might make the model a bit more realistic. This simpleness combined with the touch of randomness to give the model a more realistic vibe are definitely some strong points of using this model compared to other, more complex models. A simple model is more tractable and allows for a better understanding of what is going on in the dynamics. However, this might be a downside as well. Real-world opinion dynamics are, probably, far from simple. People don't update their opinion according to a simple probabilistic opinion dynamics model. Beside that, the cases where people have to choose between only two opinions are rather limited. Filtering algorithms, used by real social media companies, are unfortunately not publicly shared. We can however assume that they are not as simple as the ones we used in this thesis. Trying to model these far more complex systems by our very simple model, might lead to the fact that a lot of the dynamics might be missed or not well captured by our model. This downside of trying to model real-world social behavior with simple, tractable models is really a call for the urgent need for more real-world data to which the results obtained from simple models can be compared and which can help make the models capable of generating more realistic results. \\

@@

Note that for an initial $50/50$ opinion distribution none of the network models, nor the filtering algorithms could break this status quo (see Figure \ref{ev_op_50_50} and related results in reference \cite{Perra2019}). Despite this preserving of the balanced opinion distribution, the interplay of filtering algorithms and network structure can potentially shift the initial randomly distributed opinions to a more clustered opinion distribution and thus lead to polarization and formation of echo chambers.

@@

This enhancing of the formation of echo chambers around the minority opinion in case of networks with a high clustering coefficient is remarkable, especially in the case of the PR filtering algorithm. Remember from Figure \ref{ev_op_20_80} that in case of the PR method the minority opinion 0 decreases to less than $1 \%$ of the entire population for the high clustering networks. Even though the total fraction of the minority opinion strongly decreases, the high clustering forces the remaining nodes with this minority opinion to group together and form opinion clusters. The normalized value of the formation of echo chambers around the majority opinion reaches values approximately equal to $F_N(\left<P_1^{\text{nn}}\right> = 1) = 1$ when using the REC filtering algorithm. This means that the average fraction of nodes with all their neighbors carrying the majority opinion 1 remains the same before and after the time evolution. In case of the PR method this value is even slightly higher than 1, indicating a small increase in the number of majority opinion echo chambers after the time evolution.

@@

\subsection{Including a stubbornness parameter}
\label{stubbpar}
Here, we review the results of introducing stubborn people by including a stubbornness parameter $r_i$. We look at two slightly different case
\begin{itemize}
	\item Making a fraction of the nodes (going from none of the nodes to all the nodes) completely stubborn ($r_i = 1$),
	\item Making all nodes partly stubborn (going from not stubborn at all, $r = 0$, to completely stubborn, $r = 1$).
\end{itemize}
For both cases we will look at both the evolution of the opinions and the formation of echo chambers and we will do this for two different initial conditions: a balanced initial opinion distribution ($50/50$) and an unbalanced initial opinion distribution ($20/80$). \\
Once again, the opinion dynamics model described in Chapter \ref{chap4}, Section \ref{opinionmodel} is implemented with an activation probability equal to $p_{\text{act}} = 0.1$ and a maximum length of the actual time-line $S = 20$.

@@
